\documentclass[12pt]{article}
\pagenumbering{gobble}

\title{Summary of Correction in \\\textbf{Visualization of Analytic Provenance for Sensemaking}}

\date{}

\usepackage{palatino}
\usepackage[margin=1in]{geometry}

\begin{document}

\maketitle

Below are the correction points asked by the external examiner and my summary of changes in addressing them.

\begin{enumerate}
	\item Include URLs to videos in thesis.
	\\ $\rightarrow$ For each chapter, I included a demo link at the footnote of the first page of the chapter.
	
	\item Update minor language corrections throughout thesis.
	\\ $\rightarrow$ I fixed all errors marked in the two copies of the external and internal examiners. I also moved figures to top or bottom of the page as the external suggested.
	
	\item Shorten Chapter 2 by referencing relevant surveys on time-oriented data and network and tree data.
	\\ $\rightarrow$ I added references to relevant surveys and books, and shortened the text significantly in those two sections. Now, Section 2.4 and Section 2.5 are merged into one.
	
	\item Chapter 5: consider trade-off between amount of resources used and amount of data collected.
	\\ $\rightarrow$ I added a discussion in Section 5.4.8.3.
	
	\item Chapter 5: discuss a bit more about ``user engagement''. This point wasn't on the external examiner's viva notes but noted in his thesis copy.
	\\ $\rightarrow$ I added a paragraph at the end of Section 6.4.3.1 to discuss.
	
	\item Figure 6.11: discuss scalability as future work.
	\\ $\rightarrow$ The scalability of the history map was already discussed in Section 6.3.4.4 and in the second paragraph of Section 6.4.3.1. Now, I added a discussion on scalability of the knowledge map in Section 6.3.5.6.
\end{enumerate}

\end{document}