\chapter{Timeline Visualization with Set Relations of Provenance Data}

\graphicspath{{Chapter4/figures/}}
The timeline visualization in Chapter 4 cannot show events belonging to multiple sets. In this chapter, we introduce a novel timeline visualization technique, TimeSets, that helps make sense of complex temporal datasets by showing the set relationships among individual events. TimeSets visually groups events that share a topic, such as a place or a person, while preserving their temporal order. It dynamically adjusts the level of detail for each event to suit the amount of information and display estate. 

\section{Introduction}
The current limitation
\begin{itemize}
	\item no or very simple layout which is cluttered and space-inefficient
	\item designed for presenting a known story rather than interactively constructing a hidden one
\end{itemize} 

The contribution: TimeSets
\begin{itemize}
	\item clearly shows the events within a set over time and their relationships with other sets; 
	\item dynamically adjusts the level of details of each event to suit the amount of information and display estate;
	\item uses color gradient backgrounds for events belonging to multiple sets and curved set outlines to emphasize its grouping.
\end{itemize}

\section{Design}
Visual representation of individual events and sets

\section{Layout}
Describe the layout algorithm.

\section{Evaluation}
A controlled experiment with 30 participants to compare TimeSets and the state-of-the-art set visualization technique -- KelpFusion.

\section{Case Study 1: Publication Data}
Shows an application of TimeSets to publication data of 200 articles and 8 sets.

\section{Case Study 2: VAST Challenge 2014}
Shows an application of TimeSets to tweets data. 

Also use TimeSets to show user findings [not done yet! currently, findings are shown as a node-link diagram.]

\section{Conclusion}

%\input{Chapter4/sections/intro.tex}
%%\input{Chapter4/sections/related-work.tex}
%\input{Chapter4/sections/setvis.tex}
%\input{Chapter4/sections/layout.tex}
%\input{Chapter4/sections/case-study.tex}
%\input{Chapter4/sections/evaluation.tex}
%\input{Chapter4/sections/conclusion.tex}