\chapter{Making Sense of Complex Rational Relationship}
\label{chap:sensemap}

\graphicspath{{Chapter6/figures/}}

\blfootnote{Published at the \emph{IEEE Conference on Visual Analytics Science and Technology}~\cite{Nguyen2016b}.}

%\pagebreak

\autoref{chap:sensepath} introduces a timeline visualization of automatically captured user actions to enable researchers to explore rational relationship in the user's sensemaking process. In this last block of the thesis, we investigate whether and how analytic provenance can help users in their own \emph{ongoing} sensemaking processes. Users often get lost while making sense of complicated tasks using big datasets over long period. They may forget what they have done, may be unaware of where they are in the context of the overall task, and may be unsure where to continue. Visualizing the provenance of user actions to provide an overview of the sensemaking process could help address this issue. Also, during sensemaking, by allowing users to interact with their provenance information, they have opportunities to contribute additional knowledge to the automatic capture such as through grouping and linking related information. This richly semantic organization could help users express and understand the relationship of their individual findings better and remembering them more easily.

In this chapter, we introduce a visualization tool -- \emph{\textcb{SenseMap}} -- that enable users to explore \textch{complex rational relationship} of sensemaking through the visualization of such \textcm{temporal and relational} provenance information, focusing on the context of \emph{browser-based online sensemaking}. We conducted a semi-structured interview with nine participants to explore their behaviors in online sensemaking with existing browser functionality. A simplified sensemaking model based on Pirolli and Card's model is derived to better represent the behaviors we found: users iteratively \emph{collect} information sources relevant to the task, \emph{curate} them in a way that makes sense, and finally \emph{communicate} their findings to others. A series of design workshops was followed to derive requirements, discuss designs, implement and test the prototype in an agile setting. SenseMap provides multi-linked visualizations of analytic provenance and enable users to curate and communicate their findings. To explore how SenseMap is used, we conducted a user study in a naturalistic work setting with five participants completing the same sensemaking task related to their daily work activities. All participants found the visual representation and interaction of the tool intuitive to use. Three of them positively engaged with the tool and produced successful outcomes. It helped them organize information sources, quickly find and navigate to the sources they wanted, and effectively communicate their findings. 

\input{Chapter6/sections/intro.tex}
\input{Chapter6/sections/requirements.tex}
\input{Chapter6/sections/design.tex}
\input{Chapter6/sections/evaluation.tex}
\input{Chapter6/sections/conclusion.tex}