\chapter{Making Sense of Rational Relationship through User Actions}
\label{chap:sensepath}

\graphicspath{{Chapter5/figures/}}

\blfootnote{Published at the \emph{IEEE Transactions on Visualization and Computer Graphics}~\cite{Nguyen2016}.}

\pagebreak


CLEARLY MENTION WHO ARE THE USERS





\autoref{chap:schemaline} and \autoref{chap:timesets} discuss how to support users to explore complex temporal relationship of sensemaking through interactive timeline visualizations of their annotations. However, the relationship emerged in the sensemaking process could be more complicated than a temporal one. Besides understanding how events are connected, getting to know why they happened in such order is also important. This chapter investigates how to support users to explore such a rational relationship of sensemaking though visualization of provenance data. Qualitative research methods are often used to gain understanding about this kind of relationship. Such a research process is highly manual and time-consuming: researchers collect observation data, transcribe screen capture videos and think-aloud recordings, identify recurring patterns, and eventually abstract the sensemaking process into a general model. This chapter contributes a visual sensemaking tool -- \emph{\textcb{SensePath}} -- that facilitates the exploration of rational relationship in sensemaking. 

The previous two chapters visualize annotations made by users. Even though these annotations often contain high-level thinking of the users, these information resources are limitedly available. This is because making detailed and frequent notes requires a huge amount of time and effort from the users. Also, it may distract them from their sensemaking tasks. This chapter focuses on exploring \textch{rational relationship} of sensemaking through both \textcm{user annotations} and automatically captured \textcm{user actions}. SensePath provides multi-linked visualizations of the captured provenance and interactive features to support analysis. Two user-centered evaluations were conducted to explore how the tool would be used and identify any benefits it could provide. The participants found the tool intuitive and considerably reduced analysis time, helping them gain deep understanding of the sensemaking processes.

\input{Chapter5/sections/intro.tex}
\input{Chapter5/sections/related-work.tex}
\input{Chapter5/sections/requirements.tex}
\input{Chapter5/sections/design.tex}
\input{Chapter5/sections/evaluation-2.tex}
\input{Chapter5/sections/conclusion.tex}